An efficient acyclic contact planner for multiped robots

### 2017-01-05 body/intro.tex
refreshing introduction

%~ Planning legged robot locomotion remains an open issue.
Legged robots move by sequentially creating contacts with the environment while avoiding collisions. Finding the contact sequence is an instance of the motion planning problem called acyclic contact planning.

@@

To make a legged robot walk on a flat ground, efficient approaches based on the capture point~\citep{Pratt2006} can be used to choose the next contact location in a deterministic manner. Choosing the effector to create a contact with is trivial because walk follows a cyclic pattern.
Unfortunately, in the general case planning complex contact interactions is extremely challenging because the determinism disappears.
At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location). Furthermore, the contact manifold is foliated~\citep{simeon-manipulation-04}: a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this decision brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles. Lastly, the contact manifold has a null measure and is thus impossible to randomly sample. This restriction prevents the use of efficient sampling based planners to solve the problem.

For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}. Aa a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.

Obtaining reasonable computation times is critical for a contact planner to be of any practical use. Our work thus aims at breaking the complexity of the acyclic contact planning problem.  To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the non-determinism of the contact selection problem. First we introduce a low dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold. Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
This decoupling \deladp{between the planning of the path and the generation of contacts} presents drawbacks and advantages discussed in previous related literature\addadp{, which we summarize in the following}.
%~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
%~ We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
%~ of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
%~ We then use a deterministic algorithm
%~ This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
%~ This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are intrinsically not able to adapt to new situations or discover complex behaviors in unforeseen contexts.

\repadp{Acyclic contact planning}{The issue of planning acyclic contacts} was first completely described by Bretl et al. in their seminal paper  \citep{Bretl:2006:MPM:1124573.1124585}. \repadp{This problem}{The issue} requires the simultaneous handling of two \addadp{sub}problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of \deladp{acyclic} equilibrium configurations along the path. A third nontrivial problem, $\mathcal{P}_3$, 
%~ not addressed in this work, 
then consists in interpolating a complete motion between two postures of the contact sequence.  A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

Most of the papers that followed the work of \citeauthor{Bretl:2006:MPM:1124573.1124585} have explored alternative formulations to handle the combinatorial issue. Two main directions have been explored. \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence. A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}. To keep reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the computation time is far from real-time  (about 1 minute of computation for a sequence of 20 contacts).  A similar approach has been considered for manipulation by \cite{gabicciniisrr15}. \citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered. 
\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
the contacts manually for more complex interactions.
In addition to the limits of the current implementations, a major drawback of optimization-based approaches is that they only offer local convergence.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb is addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected into the contact sub-manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key \repadp{postures}{positions} in the path. \citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with 
a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
in the demonstrated scenarios. In highly-constraining cases such 
as the car egress scenario that we address here, we believe that including collision constraints in the planning is a requirement.

For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
at \gls{interactive} frame rates (we say that a \deladp{planning} method is interactive when the computation time for \repadp{planning}{one contact switch} is less than the
time \repadp{for execution}{to execute it}). While this new approach provides an exciting direction of research, currently the resulting motions
look unnatural, and do not seem applicable to real robots.

As far as robotics applications are concerned, none of the existing planners is \deladp{able to solve the problem with} \gls{interactive} \deladp{performances}.
However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}. In particular, \addadp{the algorithm of} \citeauthor{Carpentier2016} \repadp{is }{are able to achieve this with} \gls{interactive} \deladp{performances on a real robot}.
Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
The main contribution of this paper is exactly this planner.

\subsection{\deladp{Situation and} Contributions}

@@

Our solution belongs to the class of decoupled approaches, because we believe \repadp{this to be}{that the separation between the generation of the root guide path and the contact sequence is} the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. Compared to previous approaches, our solution has two main novelties:

\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
directly the contact manifold, but an approximation of the \textit{contact feasible} space, for which there exists a mapping to the contact manifold. \adnote{It is not clear what the difference is between the contact manifold and the contact feasible space}
The space is sampled thanks to the \textit{reachability condition}\deladp{ we introduce}, a low-dimensional geometric condition for \textit{contact feasibility}.

\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium. This \deladp{second step} requires the explicit computation of contact configurations\repadp{, which we guide}{. It is guided} by dedicated heuristics that quickly synthesize \repadp{feasible}{relevant} configurations.

This sequential approach is the key to the efficiency of our method, though it
can result in failures\deladp{: we show that the \textit{reachability condition} only approximates the \textit{contact feasible} space}. However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.

@@

Finally, to demonstrate the validity of our approach\deladp{ regarding the complete chain}, we show that the generated contact plans can be successfully executed, either in simulation or on the real HRP-2 robot.

>>>>

%~ Planning legged robot locomotion remains an open issue.
Legged robots move by sequentially creating contacts with the environment while avoiding collisions.
After years of research, such robots autonomously walk on flat ground, but struggle to achieve more complex motions.
To climb stairs using a handrail for instance, a humanoid robot must plan the contacts to create with its feet and hands to perform the motion safely. The issue of computing this contact sequence is a motion planning problem called acyclic contact planning.

@@

To make a legged robot walk on a flat ground, efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, choosing the effector to create a contact with is trivial because walk follows a cyclic pattern (For a humanoid the left foot always follows the right foot).
Unfortunately, in the general case planning complex contact interactions is extremely challenging:
At any given time a contact choice must be made between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location). Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling based planners to solve the problem.

For this reason previous contributions having demonstrated acyclic contact locomotion on a real robot are too computationally expensive~\citep{Bretl:2006:MPM:1124573.1124585}. As a result at the DARPA Robotics Challenge, robots relied on unsafe bipedal walking to climb stairs, instead of using the provided handrails to facilitate the motion~\citep{atkensondarpa}.

Obtaining reasonable computation times is critical for a contact planner to be of any practical use. Our work thus aims at breaking the complexity of the acyclic contact planning problem.  To do so we deal sequentially with the two main issues associated to our problem: the null measure of the contact manifold, and the combinatorial of the contact selection problem. First we introduce a low dimensional space, called the \textit{contact-feasible} space, that can be sampled and mapped efficiently to the contact manifold. Then, given a path computed in the \textit{contact-feasible} space, we propose a deterministic algorithm to generate a contact sequence along the path.
This decoupling presents drawbacks and advantages discussed in previous related literature, which we summarize in the following.
%~ whole-body trajectory and the contact location and timings, we decouple path and contact planning, as suggested by \citeauthor{Bouyarmane2009}.
%~ We first plan a path for the geometric root of the robot in a low dimensional space. We formulate this space as a computationally-efficient approximation
%~ of the \textit{equilibrium feasibile} space, that is the space of root configurations for which there exists a joint configuration such that the robot is in static equilibrium.
%~ We then use a deterministic algorithm
%~ This paper presents a geometrical approximation of this condition, and a concrete implementation of such a decoupled planner.
%~ This approximation results from a trade-off between a necessary and a sufficient condition for \textit{equilibrium feasibility}, allowing us to find solutions extremely rapidly, while preserving a high success rate in the demonstrated scenarios.

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are by definition not able to discover complex behaviors in unforeseen contexts.

The issue of planning acyclic contacts was first completely described by \citeauthor{Bretl:2006:MPM:1124573.1124585}. The issue requires the simultaneous handling of two sub-problems, $\mathcal{P}_1$: planning a relevant guide path for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: planning a discrete sequence of equilibrium configurations along the path. A third nontrivial problem, $\mathcal{P}_3$, 
%~ not addressed in this work, 
then consists in interpolating a complete motion between two postures of the contact sequence.  A key issue is to avoid combinatorial explosion when considering at the same time the possible contacts and the potential paths. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in specific situations, limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

Most of the papers that followed the work of \citeauthor{Bretl:2006:MPM:1124573.1124585} have explored alternative formulations to handle the combinatorial issue. Two main directions have been explored. \textbf{On one hand, local optimization of both the root trajectory \Pa and the contact positions $\mathcal{P}_2$} has been used, to trade the combinatorial of the complete problem for a differential complexity, at the cost of local convergence \citep{1631739}. A complete example of the potential offered by such approaches was proposed by \cite{Mordatch:2012:DCB:2185520.2185539} and successfully applied to a real robot \citep{mordatch2015}. To keep reasonable computation times, the method uses a simplified dynamic model for the avatar. Still, the computation time is far from real-time  (about 1 minute of computation for 20 contacts).  A similar approach has been considered for manipulation by \cite{gabicciniisrr15}. \citeauthor{DBLP:conf/humanoids/DeitsT14} propose to solve contact planning globally as a mixed-integer problem, but only cyclic, bipedal locomotion is considered. 
\citeauthor{dai2014whole} extend the work of \citeauthor{Posa:2014:DMT:2568343.2568352} to discover the contact sequence for landing motions, but need to specify
the contacts manually for complex interactions.
In addition to the limits of the current implementations, optimization-based approaches only converge locally.

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling is shown by \citeauthor{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$). \Pb is addressed as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution can then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. \citeauthor{Bouyarmane2009} precisely focus on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using a inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. \citeauthor{7140082} also propose a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with 
a discretized environment. This planning phase does not account for collisions, implying that replanning is required in case of failure. This approach is efficient 
in the demonstrated scenarios. In highly-constraining cases such 
as the car egress scenario that we address here, we believe that including collision constraints in the planning is a requirement.

For completeness, we lastly mention a new kind of approach, recently proposed in the computer graphics field \citep{hamalainen_cpbp_2015}. The authors
use black-box physics simulators to perform Model Predictive Control for the motion of a humanoid character, and manage to obtain dynamically consistent motions
at \gls{interactive} frame rates (we say that a method is interactive when the computation time for planning is less than the
time for execution). While this new approach provides an exciting direction of research, currently the resulting motions
look unnatural, and do not seem applicable to real robots.

As far as robotics applications are concerned, none of the existing planners is \gls{interactive}.
However, recent contributions to the interpolation between contact poses (problem $\mathcal{P}_3$) have brought promising preliminary solutions \citep{Hauser2014, herzog2015trajectory, Park116, Carpentier2016}. In particular, our algorithm proposed in \citeauthor{Carpentier2016} is \gls{interactive}.
Therefore, a planner capable of efficiently solving \Pa and \Pb could outperform all existing planners if coupled with an interpolation method solving $\mathcal{P}_3$.
The main contribution of this paper is exactly this planner.

\subsection{Contributions}

@@

Our solution belongs to the class of decoupled approaches, because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. Compared to previous approaches, our solution has two main novelties:

\noindent \textbf{Regarding $\mathcal{P}_1$}, we propose a fast guide path planning algorithm. The key to its efficiency is that it does not sample
directly the contact manifold, but an approximation of the \textit{contact feasible} space. The \textit{contact feasible} space is a low dimensional space for which there exists a mapping to the contact manifold. The space is sampled thanks to the \textit{reachability condition}, a low-dimensional geometric condition for \textit{contact feasibility}.

\noindent \textbf{Regarding $\mathcal{P}_2$},  we propose a fast method to extend a \textit{contact feasible} path into a sequence of whole-body configurations in static equilibrium. This  requires the explicit computation of contact configurations. It is guided by dedicated heuristics that quickly synthesize feasible configurations.

This sequential approach is the key to the efficiency of our method, though it
can result in failures. However, we demonstrate empirically the validity of the approach: the high success rate, combined with the low computation times, allow us to plan (and re-plan upon failure) multi-contact sequences at \gls{interactive} rates.

@@

Finally, to demonstrate the validity of our approach, we show that the generated contact plans  can be successfully executed (problem  $\mathcal{P}_3$), either in simulation or on the real HRP-2 robot.
